---
title: "Data Sources"
subtitle: "NYC Open Data Education Datasets"
format: html
execute:
  echo: true
---

# Table of Contents

1.  [NYC Open Data Overview](#org9042774)
2.  [Dataset Catalog](#orgf2f4bc3)
    1.  [Loading the Data Library](#orgc38c917)
3.  [Schools Data](#org67cbe00)
    1.  [School Directory](#org57bb58a)
    2.  [Building Space Usage](#org5588337)
4.  [Students Data](#org0cf4438)
    1.  [Demographic Snapshot](#org5e2b1d1)
    2.  [Attendance Data](#orgad55875)
5.  [Learning Outcomes Data](#org1f58851)
    1.  [Test Results](#org10b10e4)
    2.  [Graduation Data](#org1114550)
6.  [Accessing the Data](#org760926c)
    1.  [Basic Fetch](#org48e854f)
    2.  [With Filtering (SoQL)](#org487ad9d)
    3.  [Large Datasets](#org31a3c88)
7.  [Data Dictionary Reference](#orgb3c94da)
8.  [Update Schedule](#org1ddc82e)

<a id="org9042774"></a>

# NYC Open Data Overview

The [NYC Open Data Portal](https://opendata.cityofnewyork.us/) provides free public access to over 3,000 datasets from NYC agencies. The Education category contains data from the NYC Department of Education (DOE) covering schools, students, and academic outcomes.


<a id="orgf2f4bc3"></a>

# Dataset Catalog


<a id="orgc38c917"></a>

## Loading the Data Library
```{r}
## ============================================================
## Load NYC Open Data library
## ============================================================

source(here::here("src/R/setup.R"))

# Display available datasets
datasets <- list_datasets()
print(datasets)
```

<a id="org67cbe00"></a>

# Schools Data


<a id="org57bb58a"></a>

## School Directory
```{r}
## ============================================================
## School Directory Dataset
## Comprehensive list of all NYC public schools
## ============================================================

info <- get_dataset_info(NYC_DATA$school_directory)
cat("Dataset:", info$name, "\n")
cat("ID:", NYC_DATA$school_directory, "\n")
cat("Last Updated:", format(info$last_updated, "%Y-%m-%d"), "\n")
cat("Columns:", info$columns, "\n")
```

<a id="org5588337"></a>

## Building Space Usage
```{r}
## ============================================================
## Building Space Usage Dataset  
## Annual space utilization metrics
## ============================================================

info <- get_dataset_info(NYC_DATA$building_space_usage)
cat("Dataset:", info$name, "\n")
cat("ID:", NYC_DATA$building_space_usage, "\n")
cat("Description:", substr(info$description, 1, 200), "...\n")
```
<a id="org0cf4438"></a>

# Students Data


<a id="org5e2b1d1"></a>

## Demographic Snapshot
```{r}
## ============================================================
## Demographic Snapshot
## Student demographics by school
## ============================================================

info <- get_dataset_info(NYC_DATA$demographic_snapshot)
cat("Dataset:", info$name, "\n")
cat("ID:", NYC_DATA$demographic_snapshot, "\n")
cat("Use Case: Analyze student population composition\n")
```

<a id="orgad55875"></a>

## Attendance Data
```{r}
## ============================================================
## Attendance & Chronic Absenteeism
## School-level attendance metrics
## ============================================================

info <- get_dataset_info(NYC_DATA$attendance)
cat("Dataset:", info$name, "\n")
cat("ID:", NYC_DATA$attendance, "\n")
cat("Use Case: Study attendance patterns and chronic absenteeism\n")
```

<a id="org1f58851"></a>

# Learning Outcomes Data

<a id="org10b10e4"></a>

## Test Results
```{r}
## ============================================================
## State Assessment Results
## Math and ELA scores for grades 3-8
## ============================================================

cat("=== Math Test Results ===\n")
cat("ID:", NYC_DATA$math_test_results, "\n")
cat("Grades: 3-8\n")
cat("Metrics: Proficiency levels 1-4\n\n")

cat("=== ELA Test Results ===\n")
cat("ID:", NYC_DATA$ela_test_results, "\n")
cat("Grades: 3-8\n")
cat("Metrics: Proficiency levels 1-4\n")
```

<a id="org1114550"></a>

## Graduation Data
```{r}
## ============================================================
## High School Graduation Outcomes
## 4-year and 6-year cohort graduation rates
## ============================================================

info <- get_dataset_info(NYC_DATA$graduation_outcomes)
cat("Dataset:", info$name, "\n")
cat("ID:", NYC_DATA$graduation_outcomes, "\n")
cat("Metrics: Graduation rate, dropout rate, still enrolled\n")
```
<a id="org760926c"></a>

# Accessing the Data


<a id="org48e854f"></a>

## Basic Fetch
```{r}
# Fetch with automatic caching
schools <- fetch_dataset(
dataset_id = NYC_DATA$school_directory,
cache_name = "school_directory"
)
```

<a id="org487ad9d"></a>

## With Filtering (SoQL)
```{r}
# Server-side filtering
brooklyn_schools <- fetch_dataset(
  dataset_id = NYC_DATA$school_directory,
  `$where` = "boro = 'K'"  # K = Brooklyn
)
```

<a id="org31a3c88"></a>

## Large Datasets
```{r}
# Paginated fetch for large datasets
all_demographics <- fetch_large_dataset(
  dataset_id = NYC_DATA$demographic_snapshot,
  batch_size = 50000,
  cache_name = "demographics_full"
)
```

<a id="orgb3c94da"></a>

# Data Dictionary Reference

For detailed field descriptions, visit the dataset pages on NYC Open Data:

-   [Education Category](https://data.cityofnewyork.us/Education)
-   [API Documentation](https://dev.socrata.com/foundry/data.cityofnewyork.us/)

<a id="org1ddc82e"></a>

# Update Schedule

<table border="2" cellspacing="0" cellpadding="6" rules="groups" frame="hsides">

<colgroup>
<col  class="org-left" />

<col  class="org-left" />
</colgroup>
<thead>
<tr>
<th scope="col" class="org-left">Dataset</th>
<th scope="col" class="org-left">Update Frequency</th>
</tr>
</thead>
<tbody>
<tr>
<td class="org-left">School Directory</td>
<td class="org-left">Annual (August)</td>
</tr>

<tr>
<td class="org-left">Building Space</td>
<td class="org-left">Annual</td>
</tr>

<tr>
<td class="org-left">Demographics</td>
<td class="org-left">Annual</td>
</tr>

<tr>
<td class="org-left">Test Results</td>
<td class="org-left">Annual (post-testing)</td>
</tr>

<tr>
<td class="org-left">Graduation</td>
<td class="org-left">Annual (cohort-based)</td>
</tr>

<tr>
<td class="org-left">Attendance</td>
<td class="org-left">Annual</td>
</tr>
</tbody>
</table>

