
# Table of Contents

1.  [Overview](#orgf4af5d0)
2.  [Setup](#orgda166f7)
3.  [Data Loading](#org4ed7f72)
    1.  [Math Test Results](#org1674518)
    2.  [ELA Test Results](#orge190f7b)
    3.  [Graduation Outcomes](#orgda8910e)
    4.  [Regents Exam Results](#org1413fb0)
4.  [Test Score Analysis](#orgb06f684)
    1.  [Math Proficiency Trends](#org811dbad)
    2.  [Math vs ELA Comparison](#orgbb45515)
    3.  [Test Scores by Borough](#org881d3b1)
5.  [Graduation Analysis](#orgf0dd8df)
    1.  [Graduation Rate Trends](#orge4c4d39)
    2.  [Graduation by Demographics](#org2f89db8)
    3.  [Dropout Analysis](#org206e790)
6.  [Regents Exam Analysis](#org3f108b0)
    1.  [Regents Pass Rates by Subject](#org042f8a4)
7.  [Achievement Gap Analysis](#org3d56dcc)
    1.  [Performance Gap by Economic Status](#org6aab7ad)
8.  [Summary Statistics](#org1f3c748)
9.  [Key Findings](#orgcddf5fa)
    1.  [Proficiency Trends](#org189896e)
    2.  [Graduation Outcomes](#org69506b3)
    3.  [Areas for Further Investigation](#org6c7b079)

---
title: "Learning Outcomes"
subtitle: "Test Scores, Graduation Rates, and Academic Progress"
format: html
execute:
  echo: true
  warning: false
---


<a id="orgf4af5d0"></a>

# Overview

Analysis of NYC student learning outcomes including standardized test scores, graduation rates, and academic progress indicators.


<a id="orgda166f7"></a>

# Setup
```{r}
## ============================================================
## Setup: Load packages and data access functions
## ============================================================

source("src/R/setup.R")
```

<a id="org4ed7f72"></a>

# Data Loading


<a id="org1674518"></a>

## Math Test Results
```{r}
## ============================================================
## Fetch Math Test Results (Grades 3-8)
## NY State standardized assessments
## ============================================================

math_results <- fetch_dataset(
  dataset_id = NYC_DATA$math_test_results,
  cache_name = "math_results"
)

cat("Math test records:", nrow(math_results), "\n")
cat("\nColumns:\n")
print(names(math_results))
```

<a id="orge190f7b"></a>

## ELA Test Results
```{r}
## ============================================================
## Fetch ELA (English Language Arts) Test Results
## ============================================================

ela_results <- fetch_dataset(
  dataset_id = NYC_DATA$ela_test_results,
  cache_name = "ela_results"
)

cat("ELA test records:", nrow(ela_results), "\n")
```

<a id="orgda8910e"></a>

## Graduation Outcomes
```{r}
## ============================================================
## Fetch High School Graduation Outcomes
## ============================================================

graduation <- fetch_dataset(
  dataset_id = NYC_DATA$graduation_outcomes,
  cache_name = "graduation"
)

cat("Graduation records:", nrow(graduation), "\n")
glimpse(graduation)
```

<a id="org1413fb0"></a>

## Regents Exam Results
```{r}
## ============================================================
## Fetch Regents Exam Results
## NY State high school assessments
## ============================================================

regents <- fetch_dataset(
  dataset_id = NYC_DATA$regents_results,
  cache_name = "regents"
)

cat("Regents exam records:", nrow(regents), "\n")
```

<a id="orgb06f684"></a>

# Test Score Analysis


<a id="org811dbad"></a>

## Math Proficiency Trends
```{r}
## ============================================================
## Math proficiency levels over time
## Level 3+ = proficient
## ============================================================

if (exists("math_results") && nrow(math_results) > 0) {

  # Look for year and proficiency columns
  year_col <- intersect(c("year", "school_year", "test_year"), names(math_results))[1]

  if (!is.na(year_col)) {

    # Calculate proficiency rate (Level 3 + Level 4)
    math_proficiency <- math_results |>
      filter(!is.na(.data[[year_col]])) |>
      group_by(year = .data[[year_col]]) |>
          summarise(
            total_tested = sum(as.numeric(number_tested), na.rm = TRUE),
            level_3_4 = sum(as.numeric(level_3_4_count), na.rm = TRUE),
            .groups = "drop"
          ) |>
          mutate(proficiency_rate = level_3_4 / total_tested * 100) |>
          arrange(year)
    
        ggplot(math_proficiency, aes(x = year, y = proficiency_rate, group = 1)) +
          geom_line(color = "#2E86AB", linewidth = 1.2) +
          geom_point(color = "#2E86AB", size = 3) +
          geom_hline(yintercept = 50, linetype = "dashed", color = "gray50") +
          scale_y_continuous(limits = c(0, 100)) +
          labs(
            title = "NYC Math Proficiency Rate (Grades 3-8)",
            subtitle = "Percentage meeting standards (Level 3+)",
            x = "Year",
            y = "Proficiency Rate (%)"
          ) +
          theme_minimal(base_size = 14) +
          theme(
            axis.text.x = element_text(angle = 45, hjust = 1),
            plot.title = element_text(face = "bold")
          )
    
      }
    
    } else {
      cat("Math results not available\n")
    }
```

<a id="orgbb45515"></a>

## Math vs ELA Comparison
```{r}
## ============================================================
## Compare Math and ELA proficiency by year
## ============================================================

if (exists("math_results") && exists("ela_results")) {

  # Process both datasets similarly
  process_test_data <- function(df, subject) {
    year_col <- intersect(c("year", "school_year", "test_year"), names(df))[1]
    if (is.na(year_col)) return(NULL)

    df |>
      group_by(year = .data[[year_col]]) |>
      summarise(
        proficiency_rate = sum(as.numeric(level_3_4_count), na.rm = TRUE) /
                          sum(as.numeric(number_tested), na.rm = TRUE) * 100,
            .groups = "drop"
          ) |>
          mutate(subject = subject)
      }
    
      math_summary <- process_test_data(math_results, "Math")
      ela_summary <- process_test_data(ela_results, "ELA")
    
      if (!is.null(math_summary) && !is.null(ela_summary)) {
        combined <- bind_rows(math_summary, ela_summary)
    
        ggplot(combined, aes(x = year, y = proficiency_rate, color = subject, group = subject)) +
          geom_line(linewidth = 1.2) +
          geom_point(size = 3) +
          scale_color_manual(values = c("Math" = "#2E86AB", "ELA" = "#A23B72")) +
          labs(
            title = "Math vs ELA Proficiency Rates",
            subtitle = "NYC Grades 3-8 State Assessments",
            x = "Year",
            y = "Proficiency Rate (%)",
            color = "Subject"
          ) +
          theme_minimal(base_size = 14) +
          theme(
            axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "top"
          )
      }
    
    }
```

<a id="org881d3b1"></a>

## Test Scores by Borough
```{r}
## ============================================================
## Test proficiency by borough
## ============================================================

if (exists("math_results") && "borough" %in% names(math_results)) {

  boro_math <- math_results |>
    group_by(borough) |>
        summarise(
          proficiency_rate = sum(as.numeric(level_3_4_count), na.rm = TRUE) /
                            sum(as.numeric(number_tested), na.rm = TRUE) * 100,
          n_schools = n_distinct(dbn),
          .groups = "drop"
        ) |>
        filter(!is.na(borough)) |>
        mutate(borough = fct_reorder(borough, proficiency_rate))
    
      ggplot(boro_math, aes(x = proficiency_rate, y = borough, fill = borough)) +
        geom_col(show.legend = FALSE) +
        geom_text(aes(label = sprintf("%.1f%%", proficiency_rate)), 
                  hjust = -0.1, size = 4) +
        scale_fill_viridis_d(option = "mako", begin = 0.2, end = 0.8) +
        labs(
          title = "Math Proficiency by Borough",
          x = "Proficiency Rate (%)",
          y = NULL
        ) +
        theme_minimal(base_size = 14) +
        expand_limits(x = max(boro_math$proficiency_rate) * 1.15)
    
    }
```

<a id="orgf0dd8df"></a>

# Graduation Analysis


<a id="orge4c4d39"></a>

## Graduation Rate Trends
```{r}
## ============================================================
## High school graduation rate trends
## ============================================================

if (exists("graduation") && nrow(graduation) > 0) {
    
      grad_year_col <- intersect(c("cohort_year", "cohort", "year"), names(graduation))[1]
      grad_rate_col <- intersect(c("total_grads_pct", "graduation_rate", "grad_rate"), names(graduation))[1]
    
      if (!is.na(grad_year_col) && !is.na(grad_rate_col)) {
    
        grad_trend <- graduation |>
          group_by(year = .data[[grad_year_col]]) |>
          summarise(
            avg_grad_rate = mean(as.numeric(.data[[grad_rate_col]]), na.rm = TRUE),
            .groups = "drop"
          ) |>
          filter(!is.na(year), !is.na(avg_grad_rate)) |>
          arrange(year)
    
        ggplot(grad_trend, aes(x = year, y = avg_grad_rate, group = 1)) +
          geom_line(color = "forestgreen", linewidth = 1.2) +
          geom_point(color = "forestgreen", size = 3) +
          geom_hline(yintercept = 80, linetype = "dashed", color = "gray50") +
          annotate("text", x = min(grad_trend$year), y = 82, 
                   label = "80% Goal", hjust = 0, color = "gray50") +
          scale_y_continuous(limits = c(50, 100)) +
          labs(
            title = "NYC High School Graduation Rate",
            subtitle = "4-Year Cohort Graduation Rate",
            x = "Cohort Year",
            y = "Graduation Rate (%)"
          ) +
          theme_minimal(base_size = 14) +
          theme(axis.text.x = element_text(angle = 45, hjust = 1))
    
      }
    
    } else {
      cat("Graduation data not available\n")
    }
```

<a id="org2f89db8"></a>

## Graduation by Demographics
```{r}
## ============================================================
## Graduation rates by demographic group
## ============================================================

if (exists("graduation")) {
    
      # Check for demographic columns
      demo_cols <- c("asian", "black", "hispanic", "white", "male", "female",
                     "ell", "swd", "poverty")
      available_demo <- intersect(demo_cols, names(graduation))
    
      if (length(available_demo) > 0) {
        cat("Graduation Rate by Demographics:\n\n")
    
        for (col in available_demo) {
          avg_rate <- mean(as.numeric(graduation[[col]]), na.rm = TRUE)
          if (!is.na(avg_rate)) {
            cat(sprintf("  %s: %.1f%%\n", str_to_title(col), avg_rate))
          }
        }
      }
    
    }
```

<a id="org206e790"></a>

## Dropout Analysis
```{r}
## ============================================================
## Dropout rate analysis
## ============================================================

if (exists("graduation") && "dropout_pct" %in% names(graduation)) {
    
      dropout_dist <- graduation |>
        filter(!is.na(dropout_pct)) |>
        mutate(dropout_rate = as.numeric(dropout_pct))
    
      ggplot(dropout_dist, aes(x = dropout_rate)) +
        geom_histogram(bins = 25, fill = "firebrick", color = "white", alpha = 0.8) +
        geom_vline(aes(xintercept = mean(dropout_rate, na.rm = TRUE)),
                   color = "darkred", linetype = "dashed", linewidth = 1) +
        labs(
          title = "Distribution of Dropout Rates",
          subtitle = "NYC High Schools (dashed line = mean)",
          x = "Dropout Rate (%)",
          y = "Number of Schools"
        ) +
        theme_minimal(base_size = 12)
    
    }
```

<a id="org3f108b0"></a>

# Regents Exam Analysis


<a id="org042f8a4"></a>

## Regents Pass Rates by Subject
```{r}
## ============================================================
## Regents exam pass rates by subject
## ============================================================

if (exists("regents") && nrow(regents) > 0) {
    
      if ("regents_exam" %in% names(regents) || "subject" %in% names(regents)) {
    
        subject_col <- if ("regents_exam" %in% names(regents)) "regents_exam" else "subject"
    
        regents_summary <- regents |>
          group_by(subject = .data[[subject_col]]) |>
          summarise(
            total_tested = sum(as.numeric(total_tested), na.rm = TRUE),
            total_passed = sum(as.numeric(total_passed_65), na.rm = TRUE),
            .groups = "drop"
          ) |>
          mutate(
            pass_rate = total_passed / total_tested * 100,
            subject = fct_reorder(subject, pass_rate)
          ) |>
          filter(!is.na(subject), total_tested > 1000)
    
        ggplot(regents_summary, aes(x = pass_rate, y = subject)) +
          geom_col(fill = "slateblue", alpha = 0.8) +
          geom_vline(xintercept = 65, linetype = "dashed", color = "gray50") +
          labs(
            title = "Regents Exam Pass Rates by Subject",
            subtitle = "Passing score = 65+",
            x = "Pass Rate (%)",
            y = NULL
          ) +
          theme_minimal(base_size = 12)
    
      }
    
    } else {
      cat("Regents data not available\n")
    }

```
<a id="org3d56dcc"></a>

# Achievement Gap Analysis


<a id="org6aab7ad"></a>

## Performance Gap by Economic Status
```{r}
## ============================================================
## Achievement gap analysis
## Compare economically disadvantaged vs all students
## ============================================================

if (exists("math_results")) {
    
      # Check for economic need indicator
      econ_cols <- names(math_results)[grep("econ|poverty|free_lunch|need", 
                                             names(math_results), 
                                             ignore.case = TRUE)]
    
      if (length(econ_cols) > 0) {
        cat("Economic indicator columns found:\n")
        print(econ_cols)
    
        # Analysis would depend on actual column structure
        # Placeholder for gap visualization
      } else {
        cat("Economic indicator not found in test data\n")
        cat("Available columns:\n")
        print(names(math_results))
      }
    
    }
```

<a id="org1f3c748"></a>

# Summary Statistics
```{r}
## ============================================================
## Generate learning outcomes summary
## ============================================================

cat("\n========== Learning Outcomes Summary ==========\n\n")
    
    if (exists("math_results")) {
      cat("Math Assessment Records:", nrow(math_results), "\n")
    }
    
    if (exists("ela_results")) {
      cat("ELA Assessment Records:", nrow(ela_results), "\n")
    }
    
    if (exists("graduation")) {
      cat("Graduation Records:", nrow(graduation), "\n")
    
      avg_grad <- mean(as.numeric(graduation$total_grads_pct), na.rm = TRUE)
      if (!is.na(avg_grad)) {
        cat(sprintf("Average Graduation Rate: %.1f%%\n", avg_grad))
      }
    }
    
    if (exists("regents")) {
      cat("Regents Exam Records:", nrow(regents), "\n")
    }
    
    cat("\n--- Data Sources ---\n")
    cat("Math Tests:", NYC_DATA$math_test_results, "\n")
    cat("ELA Tests:", NYC_DATA$ela_test_results, "\n")
    cat("Graduation:", NYC_DATA$graduation_outcomes, "\n")
    cat("Regents:", NYC_DATA$regents_results, "\n")
```

<a id="orgcddf5fa"></a>

# Key Findings

> *This section will be populated after running the analysis with actual data.*


<a id="org189896e"></a>

## Proficiency Trends

-   [ ] Math proficiency trend (improving/declining)
-   [ ] ELA proficiency trend
-   [ ] Gap between subjects


<a id="org69506b3"></a>

## Graduation Outcomes

-   [ ] Overall graduation rate trend
-   [ ] Demographic disparities
-   [ ] Dropout hotspots


<a id="org6c7b079"></a>

## Areas for Further Investigation

-   [ ] Schools exceeding expectations
-   [ ] Intervention effectiveness
-   [ ] Resource allocation patterns

