---
title: "Schools Analysis"
subtitle: "NYC School Buildings, Locations, and Facilities"
format: html
execute:
  echo: true
  warning: false
---

# Table of Contents

1.  [Overview](#org0045ab9)
2.  [Setup](#orgc6560fc)
3.  [Data Loading](#org03f96ec)
    1.  [School Directory](#org84fa474)
    2.  [Building Space Usage](#orgfdf0785)
4.  [Exploratory Analysis](#orgfd765f4)
    1.  [Schools by Borough](#orgd8bb88c)
    2.  [School Types Distribution](#org00f8ff5)
    3.  [Geographic Distribution](#orgc53cc92)
5.  [Building Utilization Analysis](#org3a1ad53)
    1.  [Space Utilization by District](#orgdfa4eb0)
6.  [Summary Statistics](#orga540894)
7.  [Export Notes](#org278ce6e)


<a id="org0045ab9"></a>

# Overview

Analysis of NYC school buildings, locations, and facilities using DOE Open Data.


<a id="orgc6560fc"></a>

# Setup 
```{r}
## ============================================================
## Setup: Load packages and data access functions
## ============================================================

source(here::here("src/R/setup.R"))
```

<a id="org03f96ec"></a>

# Data Loading


<a id="org84fa474"></a>

## School Directory
```{r}
## ============================================================
## Fetch NYC School Directory
## Dataset: 2019-2020 School Locations
## ============================================================

schools <- fetch_dataset(
  dataset_id = NYC_DATA$school_directory,
  cache_name = "school_directory"
)

## Display structure
cat("Dataset dimensions:", nrow(schools), "rows x", ncol(schools), "columns\n")
cat("\nColumn names:\n")
print(names(schools))
```

<a id="orgfdf0785"></a>

## Building Space Usage
```{r}
## ============================================================
## Fetch DOE Building Space Usage
## Updated annually - tracks space utilization in school buildings
## ============================================================

building_space <- fetch_dataset(
  dataset_id = NYC_DATA$building_space_usage,
  cache_name = "building_space"
)

cat("Building space records:", nrow(building_space), "\n")
glimpse(building_space)
```

<a id="orgfd765f4"></a>

# Exploratory Analysis


<a id="orgd8bb88c"></a>

## Schools by Borough
```{r}
## ============================================================
## Count of schools by borough
## ============================================================

schools_by_boro <- schools |>
  count(borough = substr(borough_block_lot,1,1), name = "n_schools") |>
  mutate(borough = fct_reorder(borough, n_schools))

ggplot(schools_by_boro, aes(x = n_schools, y = borough, fill = borough)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n_schools), hjust = -0.2, size = 4) +
  scale_fill_viridis_d(option = "plasma", begin = 0.2, end = 0.8) +
  labs(
    title = "NYC Public Schools by Borough",
    subtitle = "Source: NYC Open Data School Directory",
    x = "Number of Schools",
    y = NULL
      ) +
      theme_minimal(base_size = 14) +
      theme(
        plot.title = element_text(face = "bold"),
        panel.grid.major.y = element_blank()
      ) +
      expand_limits(x = max(schools_by_boro$n_schools) * 1.1)
```

<a id="org00f8ff5"></a>

## School Types Distribution
``````{r}
    ## ============================================================
    ## Distribution of school types
    ## ============================================================
    
    school_types <- schools |>
      count(school_type = schooltype, name = "count") |>
      filter(!is.na(school_type)) |>
      mutate(
        pct = count / sum(count) * 100,
        school_type = fct_reorder(school_type, count)
      )
    
    ggplot(school_types, aes(x = count, y = school_type)) +
      geom_col(fill = "steelblue", alpha = 0.8) +
      geom_text(aes(label = sprintf("%.1f%%", pct)), hjust = -0.1, size = 3.5) +
      labs(
        title = "Distribution of NYC School Types",
        x = "Number of Schools",
        y = NULL
      ) +
      theme_minimal(base_size = 12) +
      expand_limits(x = max(school_types$count) * 1.15)
```

<a id="orgc53cc92"></a>

## Geographic Distribution
```{r}
## ============================================================
## Map of school locations
## Requires: sf package for spatial data
## ============================================================

if ("latitude" %in% names(schools) && "longitude" %in% names(schools)) {
    
      schools_geo <- schools |>
        filter(!is.na(latitude), !is.na(longitude)) |>
        mutate(
          lat = as.numeric(latitude),
          lon = as.numeric(longitude)
        ) |>
        filter(between(lat, 40.4, 41.0), between(lon, -74.3, -73.7))
    
      ggplot(schools_geo, aes(x = lon, y = lat, color = boro)) +
        geom_point(alpha = 0.5, size = 1) +
        scale_color_viridis_d(name = "Borough") +
        labs(
          title = "Geographic Distribution of NYC Schools",
          x = "Longitude",
          y = "Latitude"
        ) +
        theme_minimal() +
        coord_fixed(ratio = 1.2)
    
    } else {
      cat("Geographic coordinates not available in this dataset\n")
    }

```
<a id="org3a1ad53"></a>

# Building Utilization Analysis


<a id="orgdfa4eb0"></a>

## Space Utilization by District
```{r}
## ============================================================
## Analyze building space utilization
## ============================================================

if (exists("building_space") && nrow(building_space) > 0) {
    
      utilization_summary <- building_space |>
        group_by(district) |>
        summarise(
          n_buildings = n_distinct(building_id),
          total_capacity = sum(as.numeric(bldg_capacity), na.rm = TRUE),
          total_enrollment = sum(as.numeric(bldg_enrollment), na.rm = TRUE),
          .groups = "drop"
        ) |>
        mutate(
          utilization_rate = total_enrollment / total_capacity * 100
        ) |>
        arrange(desc(utilization_rate))
    
      cat("Top 10 Districts by Utilization Rate:\n")
      print(head(utilization_summary, 10))
    
    } else {
      cat("Building space data not available\n")
    }
```

<a id="orga540894"></a>

# Summary Statistics
```{r}
## ============================================================
## Generate summary statistics table
## ============================================================

cat("\n========== NYC Schools Summary ==========\n\n")

cat("Total Schools:", nrow(schools), "\n\n")

cat("By Borough:\n")
schools |>
      count(boro) |>
      arrange(desc(n)) |>
      print()
    
    cat("\nBy Grade Level:\n")
    if ("grades" %in% names(schools)) {
      schools |>
        count(grades) |>
        arrange(desc(n)) |>
        head(10) |>
        print()
    }
```

<a id="org278ce6e"></a>

# Export Notes

This file is designed to be:

1.  Edited in Emacs Org-mode with live R evaluation
2.  Exported to Quarto markdown (.qmd) via `org-export`
3.  Rendered to HTML via `quarto render`

See `Makefile` and `scripts/org-to-qmd.el` for automation.

