
# Table of Contents

1.  [Overview](#org0257ed1)
2.  [Setup](#org132cdb8)
3.  [Data Loading](#org1c31ea7)
    1.  [Demographic Snapshot](#org238e603)
    2.  [Enrollment Data](#org693669a)
    3.  [Attendance Data](#org98031f9)
4.  [Enrollment Analysis](#org6a54f14)
    1.  [Total Enrollment Trends](#org40750cf)
    2.  [Enrollment by Grade Level](#org9eb0cbd)
5.  [Demographics Analysis](#org295dbac)
    1.  [Demographic Composition](#orgfde3641)
    2.  [Demographics by Borough](#org50a7909)
6.  [Attendance Patterns](#orgf4eec5e)
    1.  [Chronic Absenteeism](#orgda01517)
    2.  [Attendance by School Type](#org8ba1535)
7.  [Summary Statistics](#org5bbce80)
8.  [Export Notes](#org5b00f30)

---
title: "Students Analysis"
subtitle: "Enrollment, Demographics, and Attendance Patterns"
format: html
execute:
  echo: true
  warning: false
---


<a id="org0257ed1"></a>

# Overview

Analysis of NYC student enrollment, demographics, and attendance using DOE Open Data.


<a id="org132cdb8"></a>

# Setup
```{r}
## ============================================================
## Setup: Load packages and data access functions
## ============================================================

source("src/R/setup.R")
source("src/R/nyc_data.R")
source("src/R/utils.R")
```

<a id="org1c31ea7"></a>

# Data Loading


<a id="org238e603"></a>

## Demographic Snapshot

```{r}
## ============================================================
    ## Fetch NYC School Demographics
    ## Dataset: Demographic snapshot by school
    ## ============================================================
    
    demographics <- fetch_dataset(
      dataset_id = NYC_DATA$demographic_snapshot,
      cache_name = "demographics"
    )
    
    cat("Demographics dataset:", nrow(demographics), "rows\n")
    cat("\nAvailable columns:\n")
    print(names(demographics))

```

<a id="org693669a"></a>

## Enrollment Data
```{r}

## ============================================================
## Fetch school enrollment data
## ============================================================

enrollment <- fetch_dataset(
  dataset_id = NYC_DATA$enrollment,
  cache_name = "enrollment"
)

cat("Enrollment records:", nrow(enrollment), "\n")
glimpse(enrollment)

```

<a id="org98031f9"></a>

## Attendance Data
```{r}

## ============================================================
## Fetch attendance and chronic absenteeism data
## ============================================================

attendance <- fetch_dataset(
  dataset_id = NYC_DATA$attendance,
  cache_name = "attendance"
)

cat("Attendance records:", nrow(attendance), "\n")

```

<a id="org6a54f14"></a>

# Enrollment Analysis


<a id="org40750cf"></a>

## Total Enrollment Trends
```{r}

## ============================================================
## Enrollment trends over time
## ============================================================

if ("year" %in% names(enrollment) || "school_year" %in% names(enrollment)) {

  year_col <- if ("year" %in% names(enrollment)) "year" else "school_year"

  enrollment_trend <- enrollment |>
        group_by(year = .data[[year_col]]) |>
        summarise(
          total_enrollment = sum(as.numeric(total_enrollment), na.rm = TRUE),
          .groups = "drop"
        ) |>
        arrange(year)
    
      ggplot(enrollment_trend, aes(x = year, y = total_enrollment, group = 1)) +
        geom_line(color = "steelblue", linewidth = 1.2) +
        geom_point(color = "steelblue", size = 3) +
        scale_y_continuous(labels = scales::comma) +
        labs(
          title = "NYC Public School Enrollment Over Time",
          subtitle = "Source: NYC DOE Open Data",
          x = "School Year",
          y = "Total Enrollment"
        ) +
        theme_minimal(base_size = 14) +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(face = "bold")
        )
    
    } else {
      cat("Year column not found in enrollment data\n")
    }
```

<a id="org9eb0cbd"></a>

## Enrollment by Grade Level
```{r}

## ============================================================
## Enrollment distribution by grade level
## ============================================================
    
# Look for grade-related columns
grade_cols <- names(enrollment)[grep("grade|pk|k_|_k$", names(enrollment), ignore.case = TRUE)]
    
if (length(grade_cols) > 0) {
  cat("Grade-related columns found:\n")
  print(grade_cols)

  # Pivot to long format for visualization
  enrollment_by_grade <- enrollment |>
    select(any_of(c("dbn", "school_name", grade_cols))) |>
    pivot_longer(
      cols = any_of(grade_cols),
      names_to = "grade",
      values_to = "count"
    ) |>
    group_by(grade) |>
    summarise(total = sum(as.numeric(count), na.rm = TRUE), .groups = "drop") |>
    filter(total > 0) |>
    mutate(grade = fct_reorder(grade, total))

  ggplot(enrollment_by_grade, aes(x = total, y = grade)) +
    geom_col(fill = "darkorange", alpha = 0.8) +
    scale_x_continuous(labels = scales::comma) +
    labs(
      title = "Total Enrollment by Grade Level",
      x = "Total Students",
      y = "Grade"
    ) +
    theme_minimal(base_size = 12)

} else {
  cat("No grade-level columns found\n")
}
```

<a id="org295dbac"></a>

# Demographics Analysis


<a id="orgfde3641"></a>

## Demographic Composition
```{r}

## ============================================================
## Overall demographic composition
## ============================================================

# Common demographic column patterns
demo_cols <- c("asian", "black", "hispanic", "white", "multi_racial", 
                   "native_american", "pacific_islander")
    
    available_demo <- names(demographics)[names(demographics) %in% demo_cols]
    
    if (length(available_demo) > 0) {
    
      demo_totals <- demographics |>
        summarise(across(all_of(available_demo), 
                         ~sum(as.numeric(.), na.rm = TRUE))) |>
        pivot_longer(everything(), names_to = "group", values_to = "count") |>
        mutate(
          pct = count / sum(count) * 100,
          group = str_to_title(str_replace_all(group, "_", " "))
        )
    
      ggplot(demo_totals, aes(x = "", y = pct, fill = group)) +
        geom_col(width = 1) +
        coord_polar(theta = "y") +
        geom_text(aes(label = sprintf("%.1f%%", pct)),
                  position = position_stack(vjust = 0.5),
                  size = 3) +
        scale_fill_viridis_d(name = "Group") +
        labs(
          title = "NYC Student Demographics",
          subtitle = "Percentage by Ethnicity"
        ) +
        theme_void(base_size = 12) +
        theme(plot.title = element_text(face = "bold", hjust = 0.5))
    
    } else {
      cat("Demographic columns not found. Available columns:\n")
      print(names(demographics))
    }
```

<a id="org50a7909"></a>

```{r}
## Demographics by Borough

## ============================================================
## Demographic breakdown by borough
## ============================================================

if ("borough" %in% names(demographics) || "boro" %in% names(demographics)) {

  boro_col <- if ("borough" %in% names(demographics)) "borough" else "boro"

  demo_by_boro <- demographics |>
    group_by(borough = .data[[boro_col]]) |>
    summarise(
      total_students = sum(as.numeric(total_enrollment), na.rm = TRUE),
      n_schools = n(),
      .groups = "drop"
    ) |>
    arrange(desc(total_students))

  cat("Student Distribution by Borough:\n\n")
  print(demo_by_boro)

} else {
  cat("Borough column not found\n")
}
```

<a id="orgf4eec5e"></a>

# Attendance Patterns


<a id="orgda01517"></a>

## Chronic Absenteeism
```{r}

## ============================================================
    ## Chronic absenteeism analysis
    ## (Students missing 10%+ of school days)
    ## ============================================================
    
    if (exists("attendance") && nrow(attendance) > 0) {
    
      # Look for absenteeism rate columns
      absent_cols <- names(attendance)[grep("absent|chronic|attend", 
                                            names(attendance), 
                                            ignore.case = TRUE)]
    
      cat("Attendance-related columns:\n")
      print(absent_cols)
    
      if ("chronic_absenteeism_rate" %in% names(attendance)) {
    
        absentee_dist <- attendance |>
          filter(!is.na(chronic_absenteeism_rate)) |>
          mutate(rate = as.numeric(chronic_absenteeism_rate))
    
        ggplot(absentee_dist, aes(x = rate)) +
          geom_histogram(bins = 30, fill = "coral", color = "white", alpha = 0.8) +
          geom_vline(aes(xintercept = mean(rate, na.rm = TRUE)), 
                     color = "darkred", linetype = "dashed", linewidth = 1) +
          labs(
            title = "Distribution of Chronic Absenteeism Rates",
            subtitle = "Dashed line = citywide average",
            x = "Chronic Absenteeism Rate (%)",
            y = "Number of Schools"
          ) +
          theme_minimal(base_size = 12)
    
      }
    
    } else {
      cat("Attendance data not loaded\n")
    }

```

<a id="org8ba1535"></a>

## Attendance by School Type

```{r}
## ============================================================
    ## Compare attendance across school types
    ## ============================================================
    
    if (exists("attendance") && "school_type" %in% names(attendance)) {
    
      attendance_by_type <- attendance |>
        group_by(school_type) |>
        summarise(
          n_schools = n(),
          avg_attendance = mean(as.numeric(attendance_rate), na.rm = TRUE),
          avg_chronic_absent = mean(as.numeric(chronic_absenteeism_rate), na.rm = TRUE),
          .groups = "drop"
        ) |>
        arrange(desc(avg_attendance))
    
      cat("Attendance Metrics by School Type:\n\n")
      print(attendance_by_type)
    
    }

```

<a id="org5bbce80"></a>

# Summary Statistics
```{r}

## ============================================================
    ## Generate student summary statistics
    ## ============================================================
    
    cat("\n========== NYC Students Summary ==========\n\n")
    
    if (exists("demographics")) {
      total <- sum(as.numeric(demographics$total_enrollment), na.rm = TRUE)
      cat("Total Students (from demographics):", scales::comma(total), "\n\n")
    }
    
    if (exists("enrollment")) {
      cat("Enrollment data years available:\n")
      year_col <- intersect(c("year", "school_year"), names(enrollment))[1]
      if (!is.na(year_col)) {
        print(unique(enrollment[[year_col]]))
      }
    }
    
    cat("\nData sources used:\n")
    cat("- demographic_snapshot:", NYC_DATA$demographic_snapshot, "\n")
    cat("- enrollment:", NYC_DATA$enrollment, "\n")
    cat("- attendance:", NYC_DATA$attendance, "\n")
```

<a id="org5b00f30"></a>

# Export Notes

This file exports to Quarto for web publishing. Use:

-   `C-c C-e m m` to export to markdown
-   Or run `make qmd` to batch convert all org files

